Latent Variable Clustering for Classification of Events in the AT-TPC

### 2020-06-12 doc/Manuscript/article.tex
Updates from Overleaf

\subsubsection{Clustering on latent spaces}

@@

As an alternative to relying on a pre-trained model, we also consider bespoke end-to-end clustering models, which are specifically trained on the AT-TPC data. {\color{cyan} [Ryan: I think everything after this point in this paragraph should either be removed or go in a discussion at the end. Because it is entirely about DCEC, which isn't one of the two main methods we are presenting and more immediately isn't MIXAE, which is what this section is about. Or, this section just needs to be presented as something slightly more general, but right now, ``mixture of autoencoders'' is in the title.]} We considered the deep convolutional embedded clustering (DCEC) by \citet{Guo2017} as well as the mixture of autoencoders (MIXAE) method introduced by \citet{Zhang}. While we were able to reproduce the authors' results on their data, the DCEC algorithm proved unable to cluster AT-TPC in our implementation. However, this provides valuable insight as the seeds of the clusters are constructed by a K-means algorithm. This insight contrasts with our positive results from applying a pre-trained model with k-means and highlights potentially significant differences in models trained on a supervised or unsupervised objective for clustering tasks in nuclear physics.

The MIXAE algorithm clusters data by pairing a set of encoder-decoder neural networks (autoencoder) with an auxiliary assignment network: each autoencoder constructs a representation of a given event while the assignment network assigns a set of cluster probabilities to the event. As noted by \citet{Zhang}, the MIXAe model assumes that the clusters are roughly similar in volume in the data and that one has some reasonable estimate on the number of clusters present {\color{cyan} [Ryan: Is this true of our data? I think if these assumptions are going to be mentioned, then we need to say that our data meets the criteria.}. The architecture is portrayed in figure \ref{fig:mixae}, tapered boxes denote a direction of compression in the network components.

@@

In the MIXAE algorithm the hyper-parameters to adjust are all the ordinary parameters that we introduced in table \ref{tab:convae_hyperparams}. In addition to those parameters we have the weighting of the loss terms: $\theta$, $\alpha$ and $\gamma$. These weighting parameters are attached to the reconstruction loss, sample entropy and batch-wise entropy respectively \cite{Zhang}. 
We focused on the tuning of the clustering hyper-parameters, and heuristically set the autoencoder hyper-parameters to a shallow $3\times3$ convolutional network.

@@

We repeated the same procedure described in the two previous paragraphs to the raw dataset. Each configuration was re-run a total of $N=10$ times to capture fluctuations in the performance before a final selection is made on the hyperparameters.

@@

Lastly we wish to further investigate if there are systematic differences between proton events that were placed in different clusters for the clean and full data. From figure \ref{fig:mixae_confmat} we see that the MIXAE algorithm creates two proton clusters for the filtered data, and places about fifty per-cent of the proton events in a cluster with the amorphous "other" events. We extract some proton events from these clusters to inspect whether systematic differences occur.

@@

From our analysis it is clear that we are a ways off from a fully automated unsupervised method for track classification. It is also clear that there is significant potential in both avenues explored in this work. 
In particular, the surprising ability of the K-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis.
Another surprising facet of the K-means clustering is its consistent performance. As shown in table \ref{tab:clstr_vgg}, the variance is zero for the performance metrics. This result indicates that the clusters are very clearly defined in the VGG16 latent space. As we see from the non-proton clusters if figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}  that does not necessarily imply that the physical signals are correspondingly clear.

Additionally, a clear disadvantage to the K-means method is that the researcher has to specify the number of clusters in the data themselves. Each experiment then has to be considered in light of possible reaction channels to determine a sensible number of clusters for this approach.

The same weakness is present in the MIXAE implementation. Additionally, while it showed better optimal performance than the K-means method, it struggles with inconsistent results. This inconsistency was also notably not evident from the unsupervised training-objectives of the model. These two factors conspire to limit its immediate applicability.

>>>>

\subsection{Clustering on latent spaces}

@@

As an alternative to relying on a pre-trained model, we also consider bespoke end-to-end clustering models, which are specifically trained on the AT-TPC data. {\color{cyan} [Ryan: I think everything after this point in this paragraph should either be removed or go in a discussion at the end. Because it is entirely about DCEC, which isn't one of the two main methods we are presenting and more immediately isn't MIXAE, which is what this section is about. Or, this section just needs to be presented as something slightly more general, but right now, ``mixture of autoencoders'' is in the title.]}

The MIXAE algorithm clusters data by pairing a set of encoder-decoder neural networks (autoencoder) with an auxiliary assignment network: each autoencoder constructs a representation of a given event while the assignment network assigns a set of cluster probabilities to the event. {\color[orange]{clear sentences about what the model does}} As noted by \citet{Zhang}, the MIXAE model assumes that the clusters are roughly similar in volume in the data and that one has some reasonable estimate on the number of clusters present {\color{cyan} [Ryan: Is this true of our data? I think if these assumptions are going to be mentioned, then we need to say that our data meets the criteria.}. The architecture is portrayed in figure \ref{fig:mixae}, tapered boxes denote a direction of compression in the network components.

@@

{\color{orange}[move to appendix]}
In the MIXAE algorithm the hyper-parameters to adjust are all the ordinary parameters that we introduced in table \ref{tab:convae_hyperparams}. In addition to those parameters we have the weighting of the loss terms: $\theta$, $\alpha$ and $\gamma$. These weighting parameters are attached to the reconstruction loss, sample entropy and batch-wise entropy respectively \cite{Zhang}. 
We focused on the tuning of the clustering hyper-parameters, and heuristically set the autoencoder hyper-parameters to a shallow $3\times3$ convolutional network.

@@

We repeated the same procedure described in the two previous paragraphs to the raw dataset. Each configuration was re-run a total of $N=10$ times to capture fluctuations in the performance before a final selection is made on the hyperparameters. In contrast to the previous two results the MIXAE algorith very rarely converges to a high performig configuration. In fact, most runs are indistinguishable from random chance clusterings.

@@

Lastly we wish to further investigate if there are systematic differences between proton events that were placed in different clusters for the clean and full data. From figure \ref{fig:mixae_confmat} we see that the MIXAE algorithm creates two proton clusters for the filtered data, and places about fifty per-cent of the proton events in a cluster with the amorphous "other" events.

\subsection{Alternative approaches}

{\color{orange}[binding sentence]}
We considered the deep convolutional embedded clustering (DCEC) by \citet{Guo2017} as well as the mixture of autoencoders (MIXAE) method introduced by \citet{Zhang}. While we were able to reproduce the authors' results on their data, the DCEC algorithm proved unable to cluster AT-TPC in our implementation. However, this provides valuable insight as the seeds of the clusters are constructed by a K-means algorithm. This insight contrasts with our positive results from applying a pre-trained model with k-means and highlights potentially significant differences in models trained on a supervised or unsupervised objective for clustering tasks in nuclear physics.

@@

From our analysis it is clear that we are a ways off from a fully automated unsupervised method for track classification. {\color{orange}[flip this to positive ]} It is also clear that there is significant potential in both avenues explored in this work. 
In particular, the surprising ability of the K-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis.
Another surprising facet of the K-means clustering is its consistent performance. As shown in table \ref{tab:clstr_vgg}, the variance is zero for the performance metrics. This result indicates that the clusters are very clearly defined in the VGG16 latent space. As we see from the non-proton clusters if figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}  that does not necessarily imply that the physical signals are correspondingly clear.

%Additionally, a clear disadvantage to the K-means method is that the researcher has to specify the number of clusters in the data themselves. Each experiment then has to be considered in light of possible reaction channels to determine a sensible number of clusters for this approach.

%The same weakness is present in the MIXAE implementation. Additionally, while it showed better optimal performance than the K-means method, it struggles with inconsistent results. This inconsistency was also notably not evident from the unsupervised training-objectives of the model. These two factors conspire to limit its immediate applicability.
